---
title: "Analysing movie dialogue networks in R"
author: "Pete Jones"
output: html_notebook
---

## Introduction  

This notebook works through the development of a basic workflow for reading in, describing, and visualising film dialogue network data in R. The notebook will
use data for the film *Frozen* (2013) as an illustrative example and aims to be entirely reproducible.  

Make sure you have the `data` directory in the same folder as this notebook 
file.  

## Reading in the data  

First, create a function `read.film` for reading in the dialogue data as 
collected in the raw data files.  

This function takes several arguments:  

* edge.file: the csv file containing the edges
* node.file: the csv file containing the node info
* eventlist: specify the name to be used for the event list object in R 
(default is `"lines"`) 
* nodelist: specify the name to be used for the node list object in R 
(default is `"chars"`) 
* adjacency: specify the name to be used for the adjacency matrix object in R 
(default is `"adj"`)
* offset: specify the position of the "speakerID" column relative to the 
default which is column 3 (this is useful if the function is used on data which
collects data using additional or fewer columns)  

The name specification options are useful if you want to have data for more 
than one film in the workspace at the same time, but if handling one film at a 
time (as in this notebook), then the defaults are fine and only the raw data 
file names need to be specified. 

```{r}
read.film <- function(edge.file, node.file, eventlist="lines", 
                      nodelist="chars", adjacency = "adj", offset=0) {
  # First, let's read in the event list and the node list
  lines <- read.csv(edge.file, sep=',', stringsAsFactors = FALSE)
  # Now, let's read in the node list and add some attributes we'll want later
  chars <- read.csv(node.file, sep=',', stringsAsFactors = FALSE)
  chars$nlines <- vector("numeric", nrow(chars))
  for (i in 1:nrow(chars)) {
    chars$nlines[i] <- length(which(lines[,(3+offset)]==i))
  }
  chars$linesin <- colSums(lines[,(4+offset):ncol(lines)])
  chars$gender <- ifelse(chars$charfem==1,"Female","Male")
  if((FALSE %in% (apply(lines, 2, class) %in% "integer"))==FALSE) {
    lines <- as.matrix(lines)
  }
  # Create adjacency matrix from the event list
  adj <- matrix(0,nrow(chars),nrow(chars))
  for (i in 1:nrow(chars)) {
    for (j in 1:nrow(chars)) {
      adj[i,j] <- length(which(lines[,3+offset]==i&lines[,j+3+offset]==1))
    }
  }
  colnames(adj) <- chars$character.name
  rownames(adj) <- chars$character.name
  
  # Assign the output objects to the global environment
  assign(eventlist, lines, pos = ".GlobalEnv")
  assign(nodelist, chars, pos = ".GlobalEnv")
  assign(adjacency, adj, pos = ".GlobalEnv")
}
```

Run the function on the Frozen directed dialogue data to read it into R.

```{r message=FALSE}
read.film(edge.file="data/frozenlines.csv", node.file="data/frozenchars.csv")
```

Check the data for identifiable data entry errors.

```{r}
# Create a function to check for data entry errors.
check.for.errors <- function(adjacency = adj, eventlist = lines, rec=4) {
  self.ties <- vector("numeric", length = nrow(adjacency))
  # Check diagonal for self-ties
  for (i in 1:nrow(adjacency)){
    if(adjacency[i,i]>0){
      self.ties[i] <- 1
    } else {
      self.ties[i] <- 0
    }
  }
  if(length(which(self.ties>0))>0){
    cat("Characters with self-ties: ", which(self.ties>0))
  } else {
    cat("No characters with self-ties found.")
  }
  cat("\n")
  # Check for empty rows (no recipients indicated)
  if(length(which(rowSums(eventlist[,rec:ncol(eventlist)])==0))>0){
    cat("Empty rows: ", which(rowSums(eventlist[,rec:ncol(eventlist)])==0))
  } else {
    cat("No empty rows found.")
  }
  # Check for other data entry errors (cell values not in c(0, 1))
  if(FALSE %in% unique(c(as.matrix(lines)[,rec:ncol(lines)])) %in% c("0","1")) {
    cat("\nData entry values: ", unique(c(as.matrix(lines)[,rec:ncol(lines)])))
  }
}

# Run the check
check.for.errors()
```

## Viewing and describing the data  

Now we can take a look at the event list:

```{r}
head(lines)
```

and the node list:

```{r message=FALSE}
knitr::kable(chars)
```

and finally the adjacency matrix:

```{r}
adj
```

We can create a couple of descriptive tables for the data, one summarising
some film-level metadata: 

```{r}
film.summary <- function(eventlist = lines, offset=0) {
  scenecount <- length(unique(eventlist[,2]))
  linecount <- length(unique(eventlist[,1]))
  charcount <- length(unique(eventlist[,3+offset]))
  filmsum <- data.frame(scenecount,linecount,charcount)
  colnames(filmsum) <- c("No. scenes","No. lines","No. characters")
  print(filmsum, row.names = FALSE)
}
```

```{r paged.print=FALSE}
film.summary()
```

and one summarising the gendered distribution of characters & dialogue:

```{r}
film.summary.gender <- function(nodelist = chars) {
  # Calculate the proportion of named speaking characters that are female
  numfemales <- length(which(nodelist$charfem==1))
  nummales <- length(which(nodelist$charfem==0))
  propfemchars <- (numfemales/(numfemales+nummales))*100
  # Calculate the proportion of lines spoken by females
  propfemlines <- (sum(nodelist$nlines[which(nodelist$charfem==1)])/
                     (sum(nodelist$nlines[which(nodelist$charfem==1)])+
                        sum(nodelist$nlines[which(nodelist$charfem==0)])))*100
  # Calculate the proportion of total recipients of dialogue that are female
  propfemlinesin <- (sum(nodelist$linesin[which(nodelist$charfem==1)])/
                       (sum(nodelist$linesin[which(nodelist$charfem==1)])+
                          sum(nodelist$linesin[which(nodelist$charfem==0)])))*100
  # Return a table summarising each
  filmsumg <- data.frame(round(propfemchars, digits=2), 
                         round(propfemlines, digits=2),
                         round(propfemlinesin, digits=2))
  colnames(filmsumg) <- c("% characters female","% lines out female",
                          "% lines in female")
  print(filmsumg, row.names=FALSE)
}
```

```{r paged.print=FALSE}
film.summary.gender()
```

## Plotting the network  

#### Packages  

To run all the code in this section smoothly, ensure you have the following 
packages installed:

```{r eval=FALSE}
install.packages("extrafont")
install.packages("igraph")
install.packages("ggplot2")
install.packages("ggraph")
install.packages("graphlayouts")
install.packages("network")
install.packages("scales")
install.packages("visNetwork")
```

#### Using ggraph  

To produce a nice network visualisation using a combination of the R packages
`ggraph`, `graphlayouts`:

```{r results="hide"}
# Create a function for plotting the network
plot.film.gg <- function(my.adj=adj, nodelist=chars,  filmtitle="") {
  library(igraph, warn.conflicts = FALSE, quietly = TRUE)
  library(ggraph, quietly = TRUE)
  library(graphlayouts, quietly = TRUE)
  library(extrafont, quietly = TRUE)
  library(scales, quietly = TRUE)
  # Create the igraph object
  g <- graph_from_adjacency_matrix(my.adj, weighted=TRUE, diag=FALSE)
  V(g)$gender <- nodelist$gender
  V(g)$name <- nodelist$character.name
  V(g)$size <- rescale(nodelist$nlines, to=c(2,15))
  g <- reorder_edges(g, "weight", desc = FALSE)
  # Create the ggraph plot
  p <-  ggraph(g, layout = "stress") +
    geom_edge_link(aes(colour=weight, 
                       start_cap=circle(node1.size*1.1, unit="pt"),
                       end_cap=circle(node2.size*1.1, unit="pt")), n=2, 
                   arrow = arrow(angle=20,length=unit(0.14, "inches"), 
                                 ends="last", type="closed")) +
    geom_node_point(aes(fill=gender), size = V(g)$size, shape=21, stroke=0.5,
                    colour="black") +
    geom_node_text(aes(label=name), size=4.5, 
                   nudge_y=rescale(V(g)$size, to=c(0.06,0.15))) +
    scale_fill_manual(values = c("Male"="#55467a","Female"="#ded649"), 
                      aesthetics = c("fill")) +
    scale_edge_color_gradient(low="grey85", high="grey25",trans="sqrt") +
    labs(title = paste0("Character interactions in ", filmtitle)) +
    scale_x_continuous(expand = c(0.1, 0.1)) +
    theme_graph() +
    theme(legend.position="none", plot.title=element_text(size=16, hjust=0.5))
  # And plot it!
  plot(p)
}
```

```{r message=FALSE, fig.width=11, fig.height=8}
plot.film.gg(filmtitle = "Frozen (2013)")
```

#### Using network  

For a simpler approach with fewer package dependencies, we could just plot
using `network::plot`:

```{r message=FALSE}
plot.film.basic <- function(nodelist=chars, my.adj=adj, filmtitle="", 
                            legend=FALSE) {
  library(network, quietly = TRUE)
  # Create network object from adjacency matrix
  filmnet <- network(my.adj)
  # Plot the network
  plot(filmnet, label=nodelist$character.name, edge.col='lightpink1', 
       vertex.cex=sqrt(nodelist$nlines)/3, label.pos=3, arrowhead.cex=0.7,
       vertex.col=ifelse(nodelist$charfem==1, "#ded649", "#55467a"), 
       main=paste0("Character interactions in ", filmtitle))
  if (legend==TRUE) {
    legend(x="right", c("Male","Female"), pch=21, y.intersp=0.8, 
           pt.bg=c('#55467a','#ded649'), pt.cex=2.5, cex=1.2, 
           bty="n", ncol=1)
  }
}
```

```{r message=FALSE, fig.width=11, fig.height=8}
plot.film.basic(filmtitle="Frozen (2013)")
```

#### Using visNetwork  

Or for something more interactive, we can use the `visNetwork` package:

```{r message=FALSE}
plot.film.html <- function(my.adj = adj, nodelist = chars, savenet = FALSE, 
                           filmtitle = "", edgelabs = FALSE, 
                           filename = "My visnet.html") {
  library(igraph, warn.conflicts = FALSE, quietly = TRUE)
  library(visNetwork, quietly = TRUE)
  g <- graph_from_adjacency_matrix(my.adj, weighted=TRUE, diag=FALSE)
  eg <- get.edgelist(g, names = FALSE)
  
  # To create 'Links'
  numchars <- nrow(nodelist)
  ew <- vector("numeric", nrow(eg))
  for (i in 1:numchars)
  {
    for (j in 1:numchars)
    {
      for (k in 1:nrow(eg))
        if (eg[k,1]==i & eg[k,2]==j){
          ew[k] <- sqrt(my.adj[i,j])/10
        } else { next } }
  }
  vislinks <- cbind.data.frame(eg, ew)
  if(edgelabs==TRUE){
    colnames(vislinks) <- c("from", "to", "title")} else {
      colnames(vislinks) <- c("from", "to", "Strength")
    }
  
  # To create 'Nodes'
  nodetitle <- vector("character", max(nodelist$characterID))
  for (c in 1:nrow(nodelist)) {
    nodetitle[c] <- paste0("Character: ", nodelist$character.name[c],
                           "<br />Gender: ", nodelist$gender[c], 
                           "<br />No. lines: ", nodelist$nlines[c], 
                           "<br />No. times spoken to: ", nodelist$linesin[c])
  }
  visnodes <- cbind.data.frame("id" = nodelist$characterID, 
                               "label" = nodelist$character.name,
                               "value" = nodelist$nlines, "title" = nodetitle,
                               "group" = nodelist$gender)
  
  #Generate the network object
  visnet <- visNetwork(nodes = visnodes, edges = vislinks,  
                       main=list(text=paste0("Character interactions in ", 
                                             filmtitle), 
                                 style="font-family:palatino;font-size:20px; 
                                  text-align:center;")) %>%
    visInteraction(hover = TRUE, hoverConnectedEdges = TRUE, 
                   multiselect = TRUE) %>%
    visGroups(groupname = "Female", color = 
                list(background="#F0D72A", border="#9A9158", 
                     highlight="#EDC825", hover="#FBD738")) %>%
    visGroups(groupname = "Male", color = 
                list(background="#075EC3", border="#172C76", 
                     highlight="#1075EA", hover="#2586F6")) %>%
    visEdges(arrows = list(to = list(enabled = TRUE, scaleFactor = 0.5)))
  if (savenet==TRUE) {
    visSave(visnet, file = filename)
  }
  visnet
}
```

```{r message=FALSE, fig.width=8, fig.height=5}
plot.film.html(filmtitle="Frozen (2013)")
```

## Creating a pipeline  

The purpose of this walkthrough is to illustrate how the functions are defined 
and what they do. A more practical workflow would involve defining the functions 
in a separate source script, so that they can be called from a more streamlined analysis pipeline for any new data.  

For example, let's say we have stored `read.film`, `check.for.errors`, `film.summary`, `film.summary.gender`, `plot.film.gg`, `plot.film.basic`, and `plot.film.html` in a script called `"film-functions.R"`. Then the entire 
analysis pipeline for any particular film we want to analyse might simply look something like this:

```{r eval=FALSE}
source("film-functions.R")
# Read in film data----
read.film("data/myfilmlines.csv", "data/myfilmedges.csv")
check.for.errors()
# Describe the film----
film.net.summary()
# Plot the network----
plot.film.basic(filmtitle = "Film title here")
plot.film.gg(filmtitle = "Film title here")
plot.film.html(filmtitle = "Film title here")
```

This way, we can reproduce every step of the analysis performed in this notebook 
for any new data using only 10 lines of code.  

## Session information  

In the interest of reproducibility, below is the version and package dependency 
info used in producing this notebook.

```{r}
sessionInfo()
```