---
title: "Reproducing the *Diana in the World of Men* paper"
author: "Pete Jones"
output: html_notebook
---

## Introduction

> "Be careful in the world of men, Diana. They don't deserve you."
> 
> --- Hippolyta, Queen of the Amazons, in *Wonder Woman*.

In this notebook I will walk through the process of reproducing the analysis in 
[my 2018 paper](https://doi.org/10.1080/14680777.2018.1510846) exploring the 
gendered character positions in *Wonder Woman*^[If you are interested in reading 
the paper but don't have access to the journal, get in touch and I can provide
you with a free eprint] from the raw dialogue data. The paper compares gendered character positions in *Wonder Woman* with two other films: *Thor* (2011) and 
*The Hunger Games* (2012). 

Make sure that the data folder is in the same directory as this notebook. The following packages will also need to be installed for the code to run smoothly: 

```{r eval=FALSE}
install.packages("extrafont")
install.packages("igraph")
install.packages("ggplot2")
install.packages("ggraph")
install.packages("graphlayouts")
install.packages("scales")
```

## *Wonder Woman*

### Reading in the *Wonder Woman* data

Building on the notebook `01-FilmPipeline.Rmd` in the parent 
directory, we will source in the functions and follow most of the steps of that 
pipeline.

```{r}
source('film-functions.R')
```

Now we can read in the raw data,

```{r}
read.film('data/wwlines.csv', 'data/wwchars.csv')
```

and check it for errors.

```{r}
check.for.errors()
```

### Is *Wonder Woman* female-led?

One of the arguments made in the paper is that the character interaction 
network data for *Wonder Woman* calls into question the idea that the film is 
'female-led'. 

We can find evidence for this claim in the node list for the film, which is 
where the figures for the number of lines spoken and received by each character 
can be found.

```{r message=FALSE}
knitr::kable(chars)
```

As we can see, Diana speaks notably less than Steve. As well as this, we can 
see that Diana is spoken to more than she speaks, while Steve speaks more than 
he is spoken to.

Is this just an anomaly (e.g. the result of a couple of scenes where he speaks 
much more than her)? In the paper, I argued that *no*, it is not an anomaly, as:

> Beginning with the first scene in which Steve speaks, Steve has more lines 
than Diana in 69.6% of all ensuing scenes in which one of the two characters 
speaks.

To show how I got that figure, I created a test to check exactly that:

```{r}
steve.speaks.more <- function(){
  # The first scene in which Steve speaks
  steve.scene.1 <- lines[min(which(lines[,3]==5)),2]
  # The first line in the first scene in which Steve speaks
  enter.steve <- min(which(lines[,2]==steve.scene.1))
  # The lines spoken by Steve or Diana:
  sdlines <- which(lines[enter.steve:nrow(lines),3] %in% c(1,5))+(enter.steve-1)
  # Get the scenes in which at least one of them speaks:
  sdscenes <- unique(lines[sdlines,2])
  # Create a vector to store whether Steve speaks more in each of these scenes
  steve.wins <- vector(mode="logical", length = length(sdscenes))
  # Find the percentage of these scenes in which Steve speaks more than Diana
  k <- 0
  for (scene in sdscenes){
    k <- k+1
    # test whether Steve's lines outnumber Diana's lines in each scene
    steve.wins[k] <- length(which(lines[,2]==scene & lines[,3]==5)) > 
      length(which(lines[,2]==scene & lines[,3]==1))
  }
  # Print the percentage which are TRUE (i.e. Steve speaks more)
  round((length(which(steve.wins)==TRUE)/length(steve.wins))*100, digits=2)
}
```

Which, when run, tells us:

```{r}
steve.speaks.more()
```

So it's clear that Steve does more of the talking than Diana. The summary table 
for the distribution of dialogue and named speaking characters by gender 
suggests that the pattern holds for males and females in general in the film:

```{r paged.print=FALSE}
film.summary.gender()
```

Although 55% of the speaking characters (11 out of 20) are female, only 43% of 
the lines are actually spoken by women. Female characters are more likely to be 
the recipients of dialogue than to speak, and the reverse is true for men. 

This illustrates another key claim in the paper: we should be aware of the limitations of knowing only the gendered distribution of speaking characters 
when thinking about the marginalisation of women's voices in Hollywood cinema. 

Yearly reports from the [Annenberg Inclusion Initiative](https://annenberg.usc.edu/research/aii#inequality) and the [Centre for the Study of Women in Television and 
Film](https://womenintvfilm.sdsu.edu/wp-content/uploads/2019/02/2018_Its_a_Mans_Celluloid_World_Report.pdf) tell us what percentage of speaking characters in 
top-grossing films are female, typically finding the figure to be around 30%. 
This is the most sophisticated systematic evidence we have on the 
marginalisation of the female voice in mainstream Hollywood narratives. 

However, the *Wonder Woman* data illustrates the limitations 
of relying on such figures, as the distributions of speaking characters and 
dialogue diverge significantly (largely due to the fact that 8 of the 11 female characters do not appear on screen at all in the final 102 minutes of the film).

This should serve as a reminder that when content analyses tell us that 30% of 
the speaking characters in a given year were female, this does not mean that 
30% of the dialogue was actually spoken by women.

The numbers so far raise questions over the idea that *Wonder Woman* is a 
"female-led" film. 

These questions are further illustrated by plotting the graph:

```{r message=FALSE, fig.width=11, fig.height=7.5}
plot.film.gg(filmtitle="Wonder Woman")
```

We can see that Steve and Diana occupy virtually the same positions in the 
character interaction system, and Diana's neighbourhood is nearly identical to Steve's. In fact...

How many characters does Diana speak to? 

```{r}
length(which(adj[1,]>0))
```  

One of these is Steve, so that leaves 16 characters in Diana's neighbourhood 
that Steve could also talk to. How many does he actually talk to?

```{r} 
sum(which(adj[5,]>0) %in% which(adj[1,]>0))
```  

Steve speaks to 14 of the 16 characters in Diana's neighbourhood that aren't 
him. Of the two he doesn't also speak to, one is Antiope, who dies in the scene 
in which Steve is introduced. 

From this perspective, it is hard to see how we could describe Diana as 
independent from Steve, when it seems that almost all of her character 
interactions need to be reinforced by Steve. Her narrative is *their* narrative.

### Power-with

Another thing which is evident from the network diagram is that, factoring in 
the fact that the Amazons do not appear in the final 102 minutes of the film, 
Diana has very few opportunities to interact with other women. Once she leaves Themyscira with Steve, she really is in the World of Men, where the only other 
women are a secretary and a disfigured chemist, both of whom work in service of 
much more central male characters. This means that there are no opportunities 
for a collective mode of empowerment in the film, wherein women work together 
and find strength in solidarity. in the end, the strength Diana finds to 
overcome Ares is not rooted in any kind of collective empowerment, it is rooted 
in the heroic sacrifice of her male sidekick Steve. Indeed, the vast majority 
of the narrative activity flows between these two characters. 

## *Thor*

We can compare the *Wonder Woman* data with the data from the 2011 Marvel 
Studios film *Thor*. There are a lot of similarities between the films which 
make for a fruitful comparison: both are superhero action film adaptations of 
iconic mainstream comic book characters; both are estimated to have had a $150 million budget and were marketed to similar audiences; the title characters of 
both films are demigods whose story is drawn from classical mythology; in the 
plot of each film, the character travels from their mythical homeland to the 
earth of mortals familiar to the audience, creating a fish-out-of-water 
narrative in the second half of each film; and both characters are assisted by 
a mortal love interest character of the opposite sex.

Let's read in the film, this time specifying names for the objects so we can 
keep everything in the same workspace.

```{r}
read.film('data/thorlines.csv', 'data/thorchars.csv', 
          eventlist="thorlines", nodelist="thorchars", adjacency="thoradj")
```

```{r}
check.for.errors(adjacency=thoradj, eventlist=thorlines)
```

Take a look at the distribution of dialogue and characters by gender:

```{r paged.print=FALSE}
film.summary.gender(thorchars)
```

Take a look at the character stats:

```{r message=FALSE}
knitr::kable(thorchars)
```

And plot the network:

```{r message=FALSE, fig.width=11, fig.height=7.5}
plot.film.gg(thoradj, thorchars, "Thor")
```

What does the *Thor* data show? Firstly, the hero-sidekick dynamic found in 
*Wonder Woman* plays out quite differently in *Thor*. Thor clearly speaks more 
lines than his sidekick Jane Foster. Moreover, Thor's neighbourhood is larger 
than and independent from that of Jane - Thor speaks to 12 characters, of whom 
Jane speaks to only 3 (Jane speaks to only 5 characters in total. Thor's 
interactions don't need to be reinforced by Jane, and the narrative is enacted 
in much broader terms than the heteroromantic dyad of Jane and Thor. Unlike 
Diana in *Wonder Woman*, Thor is the clear leader of the narrative in *Thor*.

## *The Hunger Games*

Finally, we can compare the results with the 2012 film *The Hunger Games*, 
which has attracted a lot of discussion for its gender representation.

Read in the film:

```{r}
read.film('data/thglines.csv', 'data/thgchars.csv', 
          eventlist="thglines", nodelist="thgchars", adjacency="thgadj")
```

```{r}
check.for.errors(adjacency=thgadj, eventlist=thglines)
```

Take a look at the distribution of dialogue and characters by gender:

```{r paged.print=FALSE}
film.summary.gender(thgchars)
```

Take a look at the character stats:

```{r message=FALSE}
knitr::kable(thgchars)
```


And plot the network:

```{r message=FALSE, fig.width=11, fig.height=7.5}
plot.film.gg(thgadj, thgchars, "The Hunger Games")
```

Again, The Hunger Games shows very different narrative patterns to 
Wonder Woman. Katniss speaks 81 more lines than any other character and is 
spoken to more than twice as much as anyone else. Katniss also charts her own 
way through the narrative, interacting with 12 other named speaking characters 
in the film. Of these 12, Peeta speaks only to 5, Haymitch speaks to 3, and 
Gale speaks to none. Katniss' narrative is much broader than any of these male 
characters.

Although only 43% of the named speaking characters are female (compared with 
*Wonder Woman*'s 55%), female characters speak a higher share of the lines of 
dialogue in *The Hunger Games* (45%) than in *Wonder Woman* (43%). Again, this 
is a reminder of the limitations of knowing only the proportion of speaking 
characters that are female as a metric of the vocal dimension of narrative 
marginalisation.

Finally, although Katniss is the clear vocal leader and narrative focus of 
*The Hunger Games*, her position is not one of isolation among men, leading to 
an individualised sense of empowerment-as-exceptionalism. Katniss' strength is 
rooted in a number of positive relationships -- with her sister Prim, with her 
mentor Effie, and most notably with fellow tribute Rue, with whom Katniss forms 
an alliance in the arena which aims to subvert the violence of the Games through 
wits, shared knowledge and solidarity. Strong female characters can be strong in 
ways that show the importance of female relationships to one's empowerment.

## Conclusion

The upshot of all this, as I argue in the paper, is that it seems that 
*Wonder Woman* is not quite as 'female-led' as it first appears, contrary to 
the strong claims made in the critical reception to the film. The narrative 
follows not Diana as an individual so much as it follows the Steve-Diana dyad 
(a dyad which is led by Steve). Diana's strength is rooted in Steve. The 
patterns that are revealed from looking at the interaction network data ask us 
to think a little differently about the kind of narrative offered by 
*Wonder Woman*, and the kind of empowerment its hero represents. *Thor* shows us 
how a similar story can play out through a narrative which establishes a more 
independent hero, while *The Hunger Games* shows us how a strong female 
character can feature in a narrative in ways which not only establish her as a 
clear, independent leader, but also develop her empowerment through 
relationships with other female characters.

That's about it for the material in the paper. Hopefully this notebook has: 

- provided some transparency for the analyses in the paper.
- provided an example of the character interaction network approach in action.

Check out the next notebook `02-WW-text.Rmd` for some additional bits of text analysis on *Wonder Woman* I added after the original piece was published.

#### Using the data

If you want to use the data for anything (e.g. doing your own analysis, or just 
using it as a teaching dataset), please feel free. The citation is:

Jones, Pete. 2018. "Diana in the World of Men: a character network approach to analysing gendered vocal representation in *Wonder Woman*." 
*Feminist Media Studies*, DOI: 10.1080/14680777.2018.1510846.

#### Session information

Below is the version and package dependency info used in producing this notebook.

```{r}
sessionInfo()
```